---
title: "Programs Staff Survey"
author: "Wenxin Cui"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(tidyverse)
library(scales)
```

# Import

```{r}
survey <- read.csv("programs_survey.csv", na = c("N/A", "n/a", "N/a", "n/A", ""))
```

# Clean

```{r}
# ensure valid entries
# TODO: some N/A's are not detected during importation
survey_clean <- survey |> 
  mutate(num_clients_active = ifelse(num_clients_active == "All", num_clients, num_clients_active))

# remove trailing semicolons
survey_clean <- survey_clean |>
  mutate(role = sub(";$","", role), 
         area = sub(";$","", area), 
         time_rank = sub(";$","", time_rank)) 

# separate rank values
survey_clean <- survey_clean |> 
   separate(col = time_rank, 
            into = paste0("time_", 1:6),
            sep = ";",
            remove = TRUE,
            fill = "right", 
            extra = "drop")

# separate area values
survey_clean <- survey_clean |>
  separate(col = area,
           into = paste0("area_", 1:5), 
           sep = ";", 
           remove = TRUE, 
           fill = "right", 
           extra = "drop")

# convert data types
 survey_clean <- survey_clean |>
   mutate(num_clients = as.numeric(num_clients), 
          num_clients_active = as.numeric(num_clients))
```

```{r}
write.csv(survey_clean, "programs_survey_clean.csv", row.names = FALSE)
```

# EDA

```{r}
N <- nrow(survey_clean)

# role composition
roles <- survey_clean |> 
  group_by(role) |> 
  summarize(n = n(), prop = round(n() / N, 2)) # no life coach participant

# role x mentor composition
survey_clean |> 
  group_by(role, is_mentor) |>
  summarize(n = n(), prop = round(n() / N, 2)) # have non-profit leadership and therapist mentors

# role x tenure composition
survey_clean |> 
  group_by(role, tenure) |>
  summarize(n = n(), prop = round(n() / N, 2)) 

survey_clean |> 
  group_by(role, tenure) |>
  summarize(n = n()) |> 
  pivot_wider(names_from = tenure, values_from = n) |> 
  mutate(across(everything(), ~replace_na(.x, 0)))

# motivation
survey_clean |> 
  ggplot(aes(x = motivation)) + 
  geom_bar() # professional experience is the main goal for most members
```

# Satisfaction

```{r}
# code all likert scales as numeric
survey_clean[survey_clean == "Disagree"] <- 1
survey_clean[survey_clean == "Somewhat Disagree"] <- 2
survey_clean[survey_clean == "Neutral"] <- 3
survey_clean[survey_clean == "Somewhat Agree"] <- 4
survey_clean[survey_clean == "Agree"] <- 5

survey_clean[survey_clean == "None"] <- 1
survey_clean[survey_clean == "Little"] <- 2
survey_clean[survey_clean == "Some"] <- 3
survey_clean[survey_clean == "Fair"] <- 4
survey_clean[survey_clean == "Plenty"] <- 5
```

```{r}
# abbreviate role titles to help with visualization
survey_clean <- survey_clean |> 
  mutate(role = ifelse(role == "Non-profit leadership & management member", "Nonprofit leadership", role), 
         role = ifelse(role == "Program developer & resource navigator", "Resource navigator", role))
```

Overall satisfaction:

```{r}
survey_clean$tenure <- factor(survey_clean$tenure, 
                              levels = c("1-2 months", "3-5 months", "6-11 months", "1-2 years", "More than 2 years"))

# longer-tenured staff are more satisfied across roles
# care coordinators experience a drop in satisfaction at 3-5 months 
# program developers/resource navigators experience a similar drop at 6-11 months
survey_clean |> 
    group_by(role, tenure) |> 
    summarize(s_mean = mean(as.numeric(satisfaction_5)), 
              .groups = "drop") |> 
    ggplot(aes(x = role, y = tenure, fill = s_mean)) + 
    geom_tile() + 
    scale_fill_gradient2(
      low = "#d73027", 
      mid = "#fee08b",      
      high = "#1a9850",    
      midpoint = 3,
      name = "Mean Satisfaction",
      limits = c(1, 5),
      breaks = c(1, 2, 3, 4, 5),
      labels = c("1 (Low)", "2 (Somewhat Low)", "3 (Neutral)", "4 (Somewhat High)", "5 (High)")
    ) + 
    labs(title = "Staff Satisfaction by Role and Tenure", 
         x = "Role", y = "Tenure") + 
    theme_minimal() # blank means no data
```

Satisfaction in terms of personal alignment:

```{r}
# in terms of skills utilization,
# care coordinators experience decrease in satisfaction at 6-11 months
# non-profit leadership does not experience change in satisfaction
# program developer/resource navigators seem to experience a slight increase in satisfaction at 3-5 months but that does not last long
# satisfaction does not change for therapists
survey_clean |> 
    group_by(role, tenure) |> 
    summarize(s_mean = mean(as.numeric(satisfaction_2)), 
              .groups = "drop") |> 
    ggplot(aes(x = role, y = tenure, fill = s_mean)) + 
    geom_tile() + 
    scale_fill_gradient2(
      low = "#d73027", 
      mid = "#fee08b",      
      high = "#1a9850",    
      midpoint = 3,
      name = "Mean Satisfaction",
      limits = c(1, 5),
      breaks = c(1, 2, 3, 4, 5),
      labels = c("1 (Low)", "2 (Somewhat Low)", "3 (Neutral)", "4 (Somewhat High)", "5 (High)")
    ) + 
    labs(title = "Staff Satisfaction by Role and Tenure", 
         x = "Role", y = "Tenure") + 
    theme_minimal()

# in terms of personal goals,
# non-profit leadership does not experience an increase in satisfaction with longer tenure
# program developer/resource navigators seem to experience a slight increase in satisfaction at 3-5 months but that does not last long (similar to skills utilization)
# therapists increase a slight increase
survey_clean |> 
    group_by(role, tenure) |> 
    summarize(s_mean = mean(as.numeric(satisfaction_3)), 
              .groups = "drop") |> 
    ggplot(aes(x = role, y = tenure, fill = s_mean)) + 
    geom_tile() + 
    scale_fill_gradient2(
      low = "#d73027", # keep this color scheme consistent throughout analyses
      mid = "#fee08b",      
      high = "#1a9850",    
      midpoint = 3,
      name = "Mean Satisfaction",
      limits = c(1, 5),
      breaks = c(1, 2, 3, 4, 5),
      labels = c("1 (Low)", "2 (Somewhat Low)", "3 (Neutral)", "4 (Somewhat High)", "5 (High)")
    ) + 
    labs(title = "Staff Satisfaction by Role and Tenure", 
         x = "Role", y = "Tenure") + 
    theme_minimal()
```

# Tenured-Specific Satisfaction Declines 

Why do care coordinators at 3-5 months report lower satisfaction than those at earlier tenure points?
Why do program developers/resource navigators at 6-11 months report lower satisfaction than those at 3-5 months?

The following analysis uses overall satisfaction as the metric. Satisfaction in terms of skill utilization and personal goals will be used to contextualize the findings or as sanity checks.

```{r}
# data are cross-sectional, so avoid talking in terms of change

# TODO: for each role that experiences satisfaction declines at some tenure point, 
# identify satisfaction domains that distinguish lower-satisfaction tenure periods from others (across tenures)
# identify support systems that are the most problematic for staff at these tenure periods (within tenure)
```

## Care Coordinators

Gaps in the training domain: 

```{r}
training_cols <- paste0("training_", c(1:7))

training_gaps_cc <- survey_clean |> 
  filter(role == "CC") |> 
  mutate(tenure_group = ifelse(tenure == "3-5 months", "low_satisfaction_period", "other_periods")) |> 
  dplyr::select(role, tenure_group, all_of(training_cols)) |> 
  pivot_longer(cols = all_of(training_cols), names_to = "training_area", values_to = "adequacy") |> 
  group_by(role, tenure_group, training_area) |>
  summarize(mean_adequacy = mean(as.numeric(adequacy), na.rm = TRUE), # remember to convert likert scale to numeric
            .groups = "drop") |> 
  pivot_wider(names_from = tenure_group, values_from = mean_adequacy) |> 
  mutate(satisfaction_diff = low_satisfaction_period - other_periods)

training_gaps_cc |> 
  mutate(satisfaction_sign = ifelse(satisfaction_diff < 0.0, "Negative", "Positive")) |> 
  ggplot(aes(x = training_area, y = satisfaction_diff, fill = satisfaction_sign)) + 
  geom_col() + 
  labs(title = "Satisfaction Gaps in Training", 
       subtitle = "Care Coordinators at 3-5 Months vs at Other Periods",
       x = "Training Area", 
       y = "Satisfaction Gap", 
       fill = "Gap Direction") + # TODO: include explanations for training codes
  scale_fill_manual(values = c("Positive" = "#1a9850", 
                               "Negative" = "#d73027")) + 
  theme_minimal() 

# care coordinators at 3-5 months feel that their training in available community resources is the most lacking compared to those at other periods, followed by established partnerships/referral networks, relevant policies, role-specific skills, client population, and lastly big picture
```

## Program Developers/Resource Navigators

Gaps in training domain: 

```{r}
training_gaps_pdrn <- survey_clean |> 
  filter(role == "PDRN") |> 
  mutate(tenure_group = ifelse(tenure == "6-11 months", "low_satisfaction_period", "other_periods")) |> 
  dplyr::select(role, tenure_group, all_of(training_cols)) |> 
  pivot_longer(cols = all_of(training_cols), names_to = "training_area", values_to = "adequacy") |> 
  group_by(role, tenure_group, training_area) |>
  summarize(mean_adequacy = mean(as.numeric(adequacy), na.rm = TRUE), # remember to convert likert scale to numeric
            .groups = "drop") |> 
  pivot_wider(names_from = tenure_group, values_from = mean_adequacy) |> 
  mutate(satisfaction_diff = low_satisfaction_period - other_periods)

training_gaps_pdrn |> 
  mutate(satisfaction_sign = ifelse(satisfaction_diff < 0.0, "Negative", "Positive")) |> 
  ggplot(aes(x = training_area, y = satisfaction_diff, fill = satisfaction_sign)) + 
  geom_col() + 
  labs(title = "Satisfaction Gaps in Training", 
       subtitle = "Resource Navigators at 6-11 Months vs at Other Periods",
       x = "Training Area", 
       y = "Satisfaction Gap", 
       fill = "Gap Direction") + # TODO: include explanations for training codes
  scale_fill_manual(values = c("Positive" = "#1a9850", 
                               "Negative" = "#d73027")) + 
  theme_minimal() 

# PDRN at 6-11 months seems to feel the most lacking in role-specific skills
```

# Client Service

Are people doing what they need to do?
Are people equipped with the resources to do what they need to do?

## Current State vs Expectation

Task completion:

```{r}
# overall task completion
survey_clean |> 
  group_by(completion_freq) |> 
  summarize(n = n()) |> 
  ggplot(aes(x = completion_freq, y = n)) + 
  geom_col() + 
  scale_y_continuous(breaks = pretty_breaks()) + 
  labs(x = "Task Completion", y = "Count") + 
  theme_minimal()

# task completion by role
survey_clean |> 
  group_by(role, completion_freq) |> 
  summarize(n = n()) |> 
  pivot_wider(names_from = completion_freq, values_from = n) |> 
  mutate(across(everything(), ~ ifelse(is.na(.), 0, .))) # nonprofit leadership is a bit concerning, but keep in mind that only two nonprofit leadership took the survey
```

Time spent by activity: 

```{r}
# time distribution
survey_clean |> 
  mutate(time_1 = ifelse(time_1 == "Waiting for information, responses, or decisions", "Waiting", time_1)) |> 
  ggplot(aes(time_1)) + 
  geom_bar() + 
  scale_y_continuous(breaks = pretty_breaks()) # nearly 30% (6 out of 21) spend the most time in routine meetings

survey_clean |> 
  mutate(time_2 = ifelse(time_2 == "Waiting for information, responses, or decisions", "Waiting", time_2)) |> 
  ggplot(aes(time_2)) + 
  geom_bar() + 
  scale_y_continuous(breaks = pretty_breaks()) # routine meetings and administrative tasks take up a considerable amount of time for nearly half of the staff members

survey_clean |> 
  filter(time_1 == "Routine meetings") |> 
  group_by(role) |> 
  summarize(n = n()) # among those who spend the most time in routine meetings, resource navigators take up the majority
```

Hours at Thaddeus: 

```{r}
# number of hours at work
survey_clean |> 
  ggplot(aes(hours_thaddeus)) + 
  geom_bar() + 
  scale_y_continuous(breaks = pretty_breaks()) + 
  theme_minimal()

survey_clean |> 
  group_by(role, hours_thaddeus) |> 
  summarize(n = n()) |> 
  pivot_wider(names_from = hours_thaddeus, values_from = n) |> 
  mutate(across(everything(), ~ ifelse(is.na(.), 0, .))) # half of the therapists work undertime (not 15-20)
                                                         # care coordinators mostly work undertime (not 20+)
                                                         # same case with case managers (15) and resource navigators (15)
```

Client responsibility: 

```{r}
# number of clients
survey_clean |> 
  filter(role %in% c("Care coordinator", "Case manager", "Therapist")) |> 
  group_by(role) |> 
  select(role, num_clients, num_clients_active) |> 
  summarize(n_clients = round(mean(num_clients, na.rm = TRUE)), 
            n_clients_active = round(mean(num_clients_active, na.rm = TRUE))) # each CC is responsible for way more clients

# how does this unbalanced distribution reflect in their perception of workload?
survey_clean |> 
  filter(role %in% c("Care coordinator", "Case manager", "Therapist")) |> 
  group_by(role, tenure) |> 
  summarize(n_clients = round(mean(num_clients, na.rm = TRUE))) # newer CC seem to handle more clients

survey_clean$tenure <- factor(survey_clean$tenure, 
                              levels = c("1-2 months", "3-5 months", "6-11 months", "1-2 years", "More than 2 years"))

survey_clean |> 
  #filter(role %in% c("Care coordinator", "Case manager", "Therapist")) |> 
  group_by(role, tenure) |> 
  summarize(workload_mean = mean(as.numeric(workload_2)), # workload fairness
            .groups = "drop") |> 
  ggplot(aes(x = role, y = tenure, fill = workload_mean)) + 
  geom_tile() +
  scale_fill_gradient2(
    low = "#d73027", 
    mid = "#fee08b",      
    high = "#1a9850",    
    midpoint = 3,
    name = "Average Workload",
    limits = c(1, 5),
    breaks = c(1, 2, 3, 4, 5),
    labels = c("Unbalanced", "Somewhat Unbalanced", "Neutral", "Somewhat Balanced", "Very Balanced")
  ) + 
  labs(title = "Workload Relative to Colleagues' Workload") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # staff generally feel that tasks are distributed fairly; care coordinators, nonprofit leadership, resource navigators, and therapists who are 3-5 months into their roles feel less that way

survey_clean |> 
  group_by(role) |> 
  summarize(workload_mean = mean(as.numeric(workload_2))) # role x tenure offers a more granular view

survey_clean |> 
  #filter(role %in% c("Care coordinator", "Case manager", "Therapist")) |> 
  group_by(role, tenure) |> 
  summarize(workload_mean = mean(as.numeric(workload_1)), # manageable
            .groups = "drop") |> 
  ggplot(aes(x = role, y = tenure, fill = workload_mean)) + 
  geom_tile() +
  scale_fill_gradient2(
    low = "#d73027", 
    mid = "#fee08b",      
    high = "#1a9850",    
    midpoint = 3,
    name = "Average Workload",
    limits = c(1, 5),
    breaks = c(1, 2, 3, 4, 5),
    labels = c("Low", "Somewhat Low", "Neutral", "Somewhat High", "High")
  ) + 
  labs(title = "Manageability") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # case managers 1-2 years into their role feel their workload is less manageable compared to colleagues who have been there shorter (mentor vs intern?); reverse pattern is also observed among care coordinators (3-5 months vs 1-2 months)

survey_clean |> 
  filter(role == "Case manager") |>
  pull(is_mentor) |> 
  unique() # confirmed: all case manager participants are interns

survey_clean |> 
  #filter(role %in% c("Care coordinator", "Case manager", "Therapist")) |> 
  group_by(role, tenure) |> 
  summarize(workload_mean = mean(as.numeric(workload_3)), # reasonable w.r.t. role
            .groups = "drop") |> 
  ggplot(aes(x = role, y = tenure, fill = workload_mean)) + 
  geom_tile() +
  scale_fill_gradient2(
    low = "#d73027", 
    mid = "#fee08b",      
    high = "#1a9850",    
    midpoint = 3,
    name = "Average Workload",
    limits = c(1, 5),
    breaks = c(1, 2, 3, 4, 5),
    labels = c("Low", "Somewhat Low", "Neutral", "Somewhat High", "High")
  ) + 
  labs(title = "Reasonability") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # staff 3-5 months into their role generally feel that their work is reasonable to a less degree
```

## Preparedness

Training:

```{r}
survey_clean |> 
  pivot_longer(cols = paste0("training_", 1:7), names_to = "training", values_to = "adequacy") |> 
  group_by(role, training) |> 
  summarize(adequacy_mean = round(mean(as.numeric(adequacy), na.rm = TRUE), 2))

# internal tools
survey_clean |> 
  group_by(role) |> 
  summarize(mean_adequacy = mean(as.numeric(training_1), na.rm = TRUE)) # lacking for case managers and nonprofit leadership

# working with the client population
survey_clean |> 
  group_by(role) |> 
  summarize(mean_adequacy = mean(as.numeric(training_4), na.rm = TRUE)) # among those who interact with clients, training is lacking for care coordinators

# community resources
survey_clean |> 
  group_by(role) |> 
  summarize(mean_adequacy = mean(as.numeric(training_2), na.rm = TRUE)) # lacking for care coordinators, nonprofit leadership, and therapists

# parterships and referral network
survey_clean |> 
  group_by(role) |> 
  summarize(mean_adequacy = mean(as.numeric(training_3), na.rm = TRUE)) # fairly lacking for all except case managers
```

Internal support:

```{r}
survey_clean[survey_clean == "Never"] <- 1
survey_clean[survey_clean == "Rarely"] <- 2
survey_clean[survey_clean == "Sometimes"] <- 3
survey_clean[survey_clean == "Often"] <- 4
survey_clean[survey_clean == "Always"] <- 5
```

```{r}
# accessible
survey_clean |> 
  group_by(role) |> 
  summarize(mean_accessible = mean(as.numeric(support_1), na.rm = TRUE), 
            sd_accessible = sd(as.numeric(support_1), na.rm = TRUE),
            median_accessible = median(as.numeric(support_1), na.rm = TRUE), # mean or median?
            mad_accessible = mad(as.numeric(support_1), na.rm = TRUE)) # lowest among nonprofit leadership and therapists

# timeliness
survey_clean |> 
  group_by(role) |> 
  summarize(mean_accessible = mean(as.numeric(support_2), na.rm = TRUE), 
            sd_accessible = sd(as.numeric(support_2), na.rm = TRUE)) # delays are more often among care coordinators, nonprofit leadership, and therapists

# effectiveness
survey_clean |> 
  group_by(role) |> 
  summarize(mean_accessible = mean(as.numeric(support_3), na.rm = TRUE), 
            sd_accessible = sd(as.numeric(support_3), na.rm = TRUE)) # nonprofit leadership perceive guidance as less effective 

# culture that is comfortable for raising questions and concerns
survey_clean |> 
  group_by(role) |> 
  summarize(mean_accessible = mean(as.numeric(support_4), na.rm = TRUE), 
            sd_accessible = sd(as.numeric(support_4), na.rm = TRUE)) # less so for resource navigators and nonprofit leadership
```

(Synthesize relevant text responses)

```{r}
survey_clean |> 
  filter(is_mentor == "Yes") # only 3 mentors among 21 participants
```

