---
title: "Programs Department Operations"
author: "Wenxin Cui"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(scales)
```

All the following results refer to the Programs Department as of November 11, 2025

```{r}
survey_data <- read.csv("programs_survey_clean.csv")

survey_data[survey_data == "None"] <- 1
survey_data[survey_data == "Little"] <- 2
survey_data[survey_data == "Some"] <- 3
survey_data[survey_data == "Fair"] <- 4
survey_data[survey_data == "Plenty"] <- 5

survey_data[survey_data == "Never"] <- 1
survey_data[survey_data == "Rarely"] <- 2
survey_data[survey_data == "Sometimes"] <- 3
survey_data[survey_data == "Often"] <- 4
survey_data[survey_data == "Always"] <- 5

survey_data[survey_data == "Disagree"] <- 1
survey_data[survey_data == "Somewhat Disagree"] <- 2
survey_data[survey_data == "Neutral"] <- 3
survey_data[survey_data == "Somewhat Agree"] <- 4
survey_data[survey_data == "Agree"] <- 5
```

# Role Composition

```{r}
N <- nrow(survey_data)

role_composition <- survey_data |> 
  mutate(role = ifelse(role == "Non-profit leadership & management member", "Non-profit leadership", role), 
         role = ifelse(role == "Program developer & resource navigator", "Resource navigator", role)) |> 
  group_by(role) |> 
  summarize(n = n(), perc = round((n / N) * 100))

role_composition |>
  ggplot(aes(x = 1, y = perc, fill = role)) + 
  geom_col() + 
  coord_polar(theta = "y") + 
  geom_text(aes(label = paste0(n, " (", paste0(perc, "%"), ")")),
            position = position_stack(vjust = 0.5), 
            size = 3.6) + 
  theme_void() + 
  scale_fill_brewer(name = NULL, palette = "RdYlGn") # compare to pop composition
```

# Work Satisfaction

```{r}
survey_data$tenure <- factor(survey_data$tenure, 
                             levels = c("1-2 months", "3-5 months", "6-11 months", "1-2 years", "More than 2 years"))

survey_data |> 
  mutate(role = ifelse(role == "Non-profit leadership & management member", "Non-profit leadership", role), 
         role = ifelse(role == "Program developer & resource navigator", "Resource navigator", role)) |>
  group_by(role, tenure) |> 
  summarize(satisfaction_mean = mean(as.numeric(satisfaction_5)), 
            .groups = "drop") |> 
  ggplot(aes(x = role, y = tenure, fill = satisfaction_mean)) + 
  geom_tile() + 
  scale_fill_gradient2(
    low = "#d73027", 
    mid = "#fee08b",      
    high = "#1a9850",    
    midpoint = 3,
    name = NULL,
    limits = c(1, 5),
    breaks = c(1, 2, 3, 4, 5),
    labels = c("Low", "Somewhat Low", "Neutral", "Somewhat High", "High")
  ) + 
  labs(x = "Role", y = "Tenure") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

# Current Performance

```{r}
# helper function that computes unbiased sd estimate
unbiased_sd <- function(v) {
  sqrt(sum((v - mean(v, na.rm = TRUE))^2) / (length(v) - 1))
}
```

## Client load (client-facing roles)

```{r}
client_facing_roles <- survey_data |> 
  filter(!is.na(num_clients) & num_clients > 0)

# overall
# each staff takes on more than 5 clients on average
client_facing_roles |> 
  summarize(n_staff = n(), 
            mean = mean(num_clients), 
            sd = round(unbiased_sd(num_clients), 2), 
            min = min(num_clients), 
            max = max(num_clients))
# no staff handles an unusually larger number of clients compared to other staff
client_facing_roles |> 
  filter(num_clients > mean(num_clients) + 2 * unbiased_sd(num_clients))

# by role
# care coordinators have the most client load, followed by therapists and case managers
(client_load_plot <- client_facing_roles |> 
  group_by(role) |> 
  summarize(n_staff = n(), 
            mean = round(mean(num_clients), 2), 
            sd = round(unbiased_sd(num_clients), 2), 
            min = min(num_clients), 
            max = max(num_clients)) |> 
  arrange(desc(mean)))

client_load_plot |> 
  ggplot(aes(x = role, y = mean)) + 
  geom_bar(stat = "identity", fill = "#fee08b") + # keep color palette consistent
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.2, 
                position = position_dodge(0.9)) + 
  scale_y_continuous(limits = c(0, 20), 
                     breaks = breaks_pretty()) + 
  labs(x = "Role", y = "Average Number of Clients") + 
  theme_minimal()
```

## Task completion (all roles)

```{r}
task_completion <- survey_data |> 
  mutate(completion_freq = as.numeric(completion_freq))
```

```{r}
# overall
# the number of staff performing below expectation is not quite alarming
task_completion |> 
  summarize(n_staff = n(), 
            mean = round(mean(completion_freq), 2), 
            sd = round(unbiased_sd(completion_freq), 2), 
            n_below_expectation = sum(completion_freq < 4), # frequent completion is used as the threshold
            perc_below_expectation = round((n_below_expectation / n_staff), 2) * 100)

# by role
# half of the non-profit leadership team do not complete tasks as expected
(task_completion_plot <- task_completion |> 
  group_by(role) |> 
  summarize(n_staff = n(), 
            mean = round(mean(completion_freq), 2), 
            sd = round(unbiased_sd(completion_freq), 2), 
            n_below_expectation = sum(completion_freq < 4), # frequent completion is used as the threshold
            perc_below_expectation = round((n_below_expectation / n_staff), 2) * 100) |> 
  arrange(mean))

task_completion_plot |> 
  mutate(role = ifelse(role == "Non-profit leadership & management member", "Non-profit leadership", role), 
         role = ifelse(role == "Program developer & resource navigator", "Resource navigator", role)) |> 
  ggplot(aes(x = role, y = mean)) + 
  geom_bar(stat = "identity", fill = "#fee08b") + # keep color palette consistent
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.2, 
                position = position_dodge(0.9)) + 
  labs(x = "Role", y = "Average Task Completion Frequency") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

# analysis of variance tests the mean completion score across all roles
# task completion is not the same across roles
aov <- aov(completion_freq ~ role, data = task_completion)
summary(aov)[[1]][["Pr(>F)"]][1] < 0.05 # significance level is set at 0.05

# post-hoc tukey comparison shows which role pairs have substantially different mean completion scores
# largest (and statistically significant) gap in task completion is seen between resource navigators and non-profit leadership
# no significant difference between case manager and resource navigators, etc.
# marginally significant difference between case managers and care coordinators, etc. 
TukeyHSD(aov)
```

## Relationship between client load and task completion (client-facing roles)

```{r}
performance <- task_completion |> 
  filter(!is.na(num_clients) & num_clients > 0)

# overall
# little correlation between overall client load and task completion
cor(performance$num_clients, performance$completion_freq, use = "complete.obs") |> round(2)
cor.test(performance$num_clients, performance$completion_freq)$p.value < 0.05

# by role
# for both care coordinators and case managers, heavier client load is associated with better task completion
# for therapists though, heavier client load is associated with worse task completion
# care coordinators show marginally significant correlation
performance |> 
  group_by(role) |> 
  summarize(cor = round(cor(num_clients, completion_freq, use = "complete.obs"), 2), 
            pval = round(cor.test(num_clients, completion_freq)$p.value, 2), 
            significant = pval < 0.05)
```

# Current Resources

## Training adequacy

```{r}
training_cols <- paste0("training_", 1:7)

training <- survey_data |> 
  select(all_of(c("role", "tenure", "num_clients", "is_mentor", training_cols))) |> 
  mutate(across(all_of(training_cols), ~as.numeric(.)))
```

```{r}
# overall
# partnerships/referral networks is the most lacking, followed by client interactions and community resources
(training_overall_plot <- training |> 
  pivot_longer(cols = all_of(training_cols), names_to = "training_area", values_to = "adequacy") |> 
  group_by(training_area) |> 
  summarize(n = sum(!is.na(adequacy)), 
            mean = round(mean(adequacy, na.rm = TRUE), 2), 
            sd = round(sd(adequacy, na.rm = TRUE), 2), # no need to calculate unbiased estimate because sample size is sufficient
            min = min(adequacy, na.rm = TRUE), 
            max = max(adequacy, na.rm = TRUE)) |> 
  arrange(mean))

training_overall_plot |> 
  mutate(training_area = case_when(
    training_area == "training_1" ~ "Internal tools and systems", 
    training_area == "training_2" ~ "Community resources", 
    training_area == "training_3" ~ "Established partnerships", 
    training_area == "training_4" ~ "Client interactions", 
    training_area == "training_5" ~ "Relevant policies", 
    training_area == "training_6" ~ "Role-specific skills", 
    training_area == "training_7" ~ "My role in the bigger picture", 
    TRUE ~ NA_character_
  )) |> 
  ggplot(aes(x = training_area, y = mean)) + 
  geom_bar(stat = "identity", fill = "#fee08b") + # keep color palette consistent
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.2, 
                position = position_dodge(0.9)) + 
  labs(x = "Training Area", y = "Average Adequacy") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
  
# by role
# most need in partnership knowledge across the board: 
# most care coordinators lack training in partnerships/referral networks (TODO: cross reference their text/interview responses)
# most case managers lack training in internal tools
# half of the non-profit leadership team lack training in all areas
# resource navigators are better trained in most areas but need more in partnerships and client interactions (not applicable to them)
# half of the therapists lack training in community resources and partnerships
training |> 
  pivot_longer(cols = all_of(training_cols), names_to = "training_area", values_to = "adequacy") |> 
  group_by(role, training_area) |> 
  summarize(n = sum(!is.na(adequacy)), 
            mean = round(mean(adequacy, na.rm = TRUE), 2),
            n_inadequate = sum(adequacy < 4, na.rm = TRUE), 
            perc_inadequate = round(n_inadequate / n, 2) * 100) |> 
  arrange(role, desc(perc_inadequate))

# by role and training type
# non-profit leadership do not have adequate training overall
# care coordinators and therapists lack client-focused training (problematic)
# case managers lack technical training
training_composites <- training |> 
  rowwise() |> 
  mutate(mean_client = mean(c_across(contains(c("2", "3", "4", "5"))), na.rm = TRUE), 
         mean_technical = mean(c_across(contains("1")), na.rm = TRUE), 
         mean_role_specific = mean(c_across(contains(c("6", "7"))), na.rm = TRUE), 
         mean_overall = mean(c_across(all_of(training_cols)), na.rm = TRUE)) |> 
  ungroup()

training_composites |> 
  group_by(role) |> 
  summarize(mean_client = round(mean(mean_client, na.rm = TRUE), 2), 
            mean_technical = round(mean(mean_technical, na.rm = TRUE), 2), 
            mean_role_specific = round(mean(mean_role_specific, na.rm = TRUE), 2),
            mean_overall = round(mean(mean_overall, na.rm = TRUE), 2)) |> 
  arrange(mean_overall)
```

## Support Effectiveness

```{r}
support_cols <- c(
  paste0("management_", 1:4), 
  paste0("team_", 1:4),
  paste0("support_", 1:4),
  paste0("growth_", 1:4)
)

support <- survey_data |> 
  select(all_of(c("role", "tenure", "num_clients", "is_mentor", support_cols)))

# support 2 is a negative statement so we need to flip the coding (e.g., disagreeing delays means agreeing efficiency)
support <- support |> 
  mutate(support_2 = case_when(
    support_2 == 1 ~ 5, 
    support_2 == 2 ~ 4,
    support_2 == 3 ~ 3, 
    support_2 == 4 ~ 2, 
    support_2 == 5 ~ 1, 
    TRUE ~ NA_real_
  ))

support <- support |> 
  mutate(across(all_of(support_cols), ~as.numeric(.)))
```

```{r}
# overall 
# support is generally fair (in the 4-5 range)
# complaint is most commonly directed towards delays, followed by internal advancement and communication of expectations
support |> 
  pivot_longer(cols = all_of(support_cols), names_to = "support_area", values_to = "satisfaction") |> 
  group_by(support_area) |> 
  summarize(n = sum(!is.na(satisfaction)), 
            mean = round(mean(satisfaction, na.rm = TRUE), 2), 
            sd = round(sd(satisfaction, na.rm = TRUE), 2), 
            min = min(satisfaction, na.rm = TRUE), 
            max = max(satisfaction, na.rm = TRUE)) |> 
  arrange(mean)

# by role and support type
# nonprofit leadership have the lowest support overall, with particularly low support in communication and troubleshooting
# therapists face particularly low support in troubleshooting
support_composites <- support |> 
  rowwise() |> 
  mutate(mean_comm = mean(c_across(contains("management")), na.rm = TRUE), 
         mean_culture = mean(c_across(contains("team")), na.rm = TRUE), 
         mean_help = mean(c_across(contains("support")), na.rm = TRUE), 
         mean_growth = mean(c_across(contains("growth")), na.rm = TRUE), 
         mean_overall = mean(c_across(all_of(support_cols)), na.rm = TRUE)) |> 
  ungroup()

support_composites |> 
  group_by(role) |> 
  summarize(mean_comm = round(mean(mean_comm), 2), 
            mean_culture = round(mean(mean_culture), 2), 
            mean_help = round(mean(mean_help), 2), 
            mean_growth = round(mean(mean_growth), 2), 
            mean_overall = round(mean(mean_overall), 2)) |> 
  arrange(mean_overall)
```

# Relationship between Resources and Performance

Present these results first before breaking training and support systems down into types

```{r}
resource_performance <- task_completion |> 
  select(all_of(c("role", "tenure", "num_clients", "is_mentor", 
                  "completion_freq", training_cols, support_cols))) |> 
  mutate(across(all_of(training_cols), ~as.numeric(.))) |> 
  mutate(across(all_of(support_cols), ~as.numeric(.)))
```

```{r}
resource_performance_composites <- resource_performance |> 
  rowwise() |> 
  mutate(mean_training = mean(c_across(all_of(training_cols)), na.rm = TRUE), 
         mean_support = mean(c_across(all_of(support_cols)), na.rm = TRUE), 
         mean_overall = mean(c(mean_training, mean_support), na.rm = TRUE))

# overall
# client load has little association with resource or task completion
resource_performance_composites |> 
  select(num_clients, mean_training, mean_support, mean_overall, completion_freq) |> 
  cor(use = "complete.obs")

# resource has significant association with task completion
cor.test(resource_performance_composites$mean_training, 
         resource_performance_composites$completion_freq)
cor.test(resource_performance_composites$mean_support, 
         resource_performance_composites$completion_freq)
cor.test(resource_performance_composites$mean_overall, 
         resource_performance_composites$completion_freq)

resource_performance_composites |> 
  filter(role != "Non-profit leadership & management member") |> 
  group_by(role) |> 
  summarize(cor_training = cor.test(mean_training, completion_freq)$p.value, 
            cor_support = cor.test(mean_training, completion_freq)$p.value)

# client-facing roles
median_client_load <- resource_performance_composites |> 
  filter(!is.na(num_clients) & num_clients > 0) |> 
  pull(num_clients) |> 
  median() # use median here (8) because the distribution of client load is right skewed

# most staff have adequate resources and can complete tasks effectively (8 out of 11)
# typical client load and inadequate resources are associated with the poorest performance
# having adequate resources has more impact on task completion than changing client load (1.2 point increase controlling for client load vs 0.5 point increase controlling for resource adequacy)
(relationship_plot <- resource_performance_composites |> 
  filter(!is.na(num_clients) & num_clients > 0) |> 
  mutate(client_load = ifelse(num_clients > median_client_load, "high", "typical"),
         resource_adequacy = ifelse(mean_overall >= 4, "adequate", "inadequate")) |> 
  group_by(client_load, resource_adequacy) |> 
  summarize(n = n(), 
            mean_completion = round(mean(completion_freq), 2), 
            sd_completion = round(unbiased_sd(completion_freq), 2)) |> # unbiased sd is used to adjust for small sample sizes
  arrange(desc(client_load), desc(resource_adequacy)))

relationship_plot$client_load <- factor(relationship_plot$client_load, levels = c("high", "typical"))
relationship_plot$resource_adequacy <- factor(relationship_plot$resource_adequacy, levels = c("inadequate", "adequate"))

relationship_plot |> 
  ggplot(aes(x = resource_adequacy, y = client_load, fill = mean_completion)) + 
  geom_tile() + 
  scale_fill_gradient2(
    low = "#d73027", 
    mid = "#fee08b",      
    high = "#1a9850",    
    midpoint = 3,
    name = "Task Completion Frequency",
    limits = c(1, 5),
    breaks = c(1, 2, 3, 4, 5),
    labels = c("Never", "Rarely", "Sometimes", "Often", "Always")
  ) + 
  labs(x = "Resource Adequacy", y = "Client Load") + 
  theme_minimal()

resource_performance_composites |> 
  filter(!is.na(num_clients) & num_clients > 0) |> 
  mutate(resource_adequacy = ifelse(mean_overall >= 4, "adequate", "inadequate")) |> 
  filter(resource_adequacy == "inadequate") |> 
  group_by(role) |> 
  summarize(n_staff = n(), 
            mean_training_gap = round(4 - mean(mean_training, na.rm = TRUE), 2), 
            mean_support_gap = round(4 - mean(mean_support, na.rm = TRUE), 2),
            mean_performance = round(mean(completion_freq, na.rm = TRUE), 2))
```




