---
title: "Programs Staff Experience Analysis: Findings"
author: "Wenxin Cui"
output: pdf_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(scales)
```

**Disclaimer**: The findings in this analysis reflect the status of the Programs Department as of **November 11, 2025**. Since then, structural changes have been implemented within the department. Results may differ if the analysis is repeated using more recent data.

```{r load data}
survey_clean <- read.csv("programs_survey_clean.csv")
```

# Department Overview

## Role composition

```{r}
N <- nrow(survey_clean)

role_composition <- survey_clean %>%
  group_by(role) %>% 
  summarize(n = n(), perc = round((n / N) * 100))

role_composition %>%
  ggplot(aes(x = 1, y = perc, fill = role)) + 
  geom_col() + 
  coord_polar(theta = "y") + 
  geom_text(aes(label = paste0(n, " (", paste0(perc, "%"), ")")),
            position = position_stack(vjust = 0.5), 
            size = 3.6) + 
  theme_void() + 
  scale_fill_brewer(name = NULL, palette = "RdYlGn")
```

## Work satisfaction

```{r}
survey_clean$tenure <- factor(survey_clean$tenure, 
                              levels = c("1-2 months", "3-5 months", "6-11 months", "1-2 years", "More than 2 years"))

survey_clean %>% 
  group_by(role, tenure) %>%
  summarize(mean_satisfaction = mean(as.numeric(satisfaction_5)), 
            .groups = "drop") |> 
  ggplot(aes(x = role, y = tenure, fill = mean_satisfaction)) + 
  geom_tile() + 
  scale_fill_gradient2(
    low = "#d73027", 
    mid = "#fee08b",      
    high = "#1a9850",    
    midpoint = 3,
    name = NULL,
    limits = c(1, 5),
    breaks = c(1, 2, 3, 4, 5),
    labels = c("Low", "Somewhat Low", "Neutral", "Somewhat High", "High")
  ) + 
  labs(x = "Role", y = "Tenure") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

# Staff Workload and Performance

```{r helper}
# helper function computing unbiased standard deviation estimate
unbiased_sd <- function(v) {
  sqrt(sum((v - mean(v, na.rm = TRUE))^2) / (length(v) - 1))
}
```

## Client load (client-facing roles only)

```{r setup}
client_facing_roles <- survey_clean %>% 
  filter(!is.na(num_clients) & num_clients > 0)
```

```{r overall}
# each staff member typically manages more than five clients
client_facing_roles %>% 
  summarize(n_staff = n(), 
            mean = mean(num_clients), 
            sd = round(unbiased_sd(num_clients), 2), 
            min = min(num_clients), 
            max = max(num_clients))

# three staff members have notably higher client caseloads than their peers
client_facing_roles %>% 
  filter(num_clients > mean(num_clients) + unbiased_sd(num_clients))
# none of the staff members has notably lower client caseloads than their peers
client_facing_roles %>% 
  filter(num_clients < mean(num_clients) - unbiased_sd(num_clients))
```

```{r by role}
# care coordinators have the heaviest client load, followed by therapists and case managers
(client_load <- client_facing_roles %>% 
  group_by(role) %>%
  summarize(n_staff = n(), 
            mean = round(mean(num_clients), 2), 
            sd = round(unbiased_sd(num_clients), 2), 
            min = min(num_clients), 
            max = max(num_clients)) %>% 
  arrange(desc(mean)))

client_load %>% 
  ggplot(aes(x = role, y = mean)) + 
  geom_bar(stat = "identity", fill = "#fee08b") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.2, 
                position = position_dodge(0.9)) + 
  scale_y_continuous(limits = c(0, 20), 
                     breaks = breaks_pretty()) + 
  labs(x = "Role", y = "Average Number of Clients") + 
  theme_minimal()
```

## Task completion (all roles)

```{r setup}
task_completion <- survey_clean %>% 
  mutate(completion_freq = as.numeric(completion_freq))
```

```{r overall}
# 3 out of 21 staff members (14.29%) reported difficulty completing tasks as expected
task_completion %>%
  summarize(n_staff = n(), 
            mean = round(mean(completion_freq), 2), 
            sd = round(unbiased_sd(completion_freq), 2), 
            n_below_expectation = sum(completion_freq < 4), # completing tasks "often" is used as the threshold
            perc_below_expectation = round((n_below_expectation / n_staff) * 100, 2))
```

```{r by role}
# half of non-profit leadership reported difficulty completing tasks as expected
# all case managers and resource navigators reported no issue with completion
(task_completion_by_role <- task_completion %>%
  group_by(role) %>% 
  summarize(n_staff = n(), 
            mean = round(mean(completion_freq), 2), 
            sd = round(unbiased_sd(completion_freq), 2), 
            n_below_expectation = sum(completion_freq < 4), # frequent completion is used as the threshold
            perc_below_expectation = round((n_below_expectation / n_staff) * 100, 2)) |> 
  arrange(mean))

task_completion_by_role %>%
  ggplot(aes(x = role, y = mean)) + 
  geom_bar(stat = "identity", fill = "#fee08b") + 
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.2, 
                position = position_dodge(0.9)) + 
  labs(x = "Role", y = "Average Task Completion Frequency") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r anova}
# analysis of variance (F-test) compares mean task scompletion across all roles
# completion is not the same across roles
aov <- aov(completion_freq ~ role, data = task_completion)
summary(aov)[[1]][["Pr(>F)"]][1] < 0.05 # significance level is set at 0.05

# post-hoc tukey comparison shows which role pairs have substantially different mean task completion
# largest (and statistically significant) gap in completion is seen between resource navigators and non-profit leadership
# marginally significant difference between non-profit leadership and case managers, non-profit leadership and care coordinators, therapists and resource navigators
# little to no difference between other role pairs
TukeyHSD(aov)
```

## Relationship between client load and task completion (client-facing roles only)

```{r setup}
performance <- task_completion %>%
  filter(!is.na(num_clients) & num_clients > 0)
```

```{r overall}
# little correlation between overall client load and task completion
cor(performance$num_clients, performance$completion_freq, use = "complete.obs") |> round(2)
cor.test(performance$num_clients, performance$completion_freq)$p.value < 0.05
```

```{r by role}
# for care coordinators and case managers, heavier client load is associated with better completion
# for therapists though, heavier client load is associated with worse completion
# the correlation is marginally significant for care coordinators
performance %>%
  group_by(role) %>%
  summarize(cor = round(cor(num_clients, completion_freq, use = "complete.obs"), 2), 
            pval = round(cor.test(num_clients, completion_freq)$p.value, 2), 
            significant = pval < 0.05)
```

# Internal Resources

## Training

```{r setup}
training_cols <- paste0("training_", 1:7)

training <- survey_clean %>%
  select(all_of(c("role", "tenure", "num_clients", "is_mentor", training_cols))) %>% 
  mutate(across(all_of(training_cols), ~as.numeric(.)))
```

```{r overall}
# knowledge of partnerships/referral networks is the most lacking, followed by client management and community resources
(training_overall <- training %>% 
  pivot_longer(cols = all_of(training_cols), names_to = "training_area", values_to = "adequacy") %>% 
  group_by(training_area) %>%
  summarize(n = sum(!is.na(adequacy)), 
            mean = round(mean(adequacy, na.rm = TRUE), 2), 
            sd = round(sd(adequacy, na.rm = TRUE), 2), # not using unbiased estimate here because sample size is sufficient
            min = min(adequacy, na.rm = TRUE), 
            max = max(adequacy, na.rm = TRUE)) |> 
  arrange(mean))

training_overall |> 
  mutate(training_area = case_when(
    training_area == "training_1" ~ "Internal tools and systems", 
    training_area == "training_2" ~ "Community resources", 
    training_area == "training_3" ~ "Established partnerships", 
    training_area == "training_4" ~ "Client interactions", 
    training_area == "training_5" ~ "Relevant policies", 
    training_area == "training_6" ~ "Role-specific skills", 
    training_area == "training_7" ~ "My role in the bigger picture", 
    TRUE ~ NA_character_
  )) |> 
  ggplot(aes(x = training_area, y = mean)) + 
  geom_bar(stat = "identity", fill = "#fee08b") + 
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.2, 
                position = position_dodge(0.9)) + 
  labs(x = "Training Area", y = "Average Adequacy") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r by role}
# half of non-profit leadership reported lacking training in all areas
# large share of case managers reported inadequate training in internal tools
# large share of care coordinators, therapists, and resource navigators all reported lacking training in client service (e.g. referral networks, interactions with clients, relevant policies) 
training %>%
  pivot_longer(cols = all_of(training_cols), names_to = "training_area", values_to = "adequacy") |> 
  group_by(role, training_area) |> 
  summarize(n = sum(!is.na(adequacy)), 
            mean = round(mean(adequacy, na.rm = TRUE), 2),
            n_inadequate = sum(adequacy < 4, na.rm = TRUE), # "fair" amount of training is used as the threshold
            perc_inadequate = round((n_inadequate / n) * 100)) |> 
  arrange(role, desc(perc_inadequate))
```

```{r by role and training category}
# develop a composite score for each major training category by weighting all constituent areas equally

# non-profit leadership reported inadequate training across all types
# care coordinators and therapists reported lacking client-focused training
# case managers reported lacking technical training
training_composites <- training %>% 
  rowwise() %>%
  mutate(client_score = mean(c_across(contains(c("2", "3", "4", "5"))), na.rm = TRUE), 
         technical_score = mean(c_across(contains("1")), na.rm = TRUE), 
         role_specific_score = mean(c_across(contains(c("6", "7"))), na.rm = TRUE), 
         overall_score = mean(c_across(all_of(training_cols)), na.rm = TRUE)) %>% 
  ungroup()

training_composites %>% 
  group_by(role) %>% 
  summarize(mean_client = round(mean(client_score, na.rm = TRUE), 2), 
            mean_technical = round(mean(technical_score, na.rm = TRUE), 2), 
            mean_role_specific = round(mean(role_specific_score, na.rm = TRUE), 2),
            mean_overall = round(mean(overall_score, na.rm = TRUE), 2)) %>%
  arrange(mean_overall)
```

## Management and work culture (collectively referred to as "support")

```{r setup}
support_cols <- c(
  paste0("management_", 1:4), 
  paste0("team_", 1:4),
  paste0("support_", 1:4),
  paste0("growth_", 1:4)
)

support <- survey_clean %>%
  select(all_of(c("role", "tenure", "num_clients", "is_mentor", support_cols)))

# support_2 is a negative statement so flip the coding (e.g. disagreeing to delays means agreeing to efficiency)
support <- support %>%
  mutate(support_2 = case_when(
    support_2 == 1 ~ 5, 
    support_2 == 2 ~ 4,
    support_2 == 3 ~ 3, 
    support_2 == 4 ~ 2, 
    support_2 == 5 ~ 1, 
    TRUE ~ NA_real_
  ))

support <- support %>% 
  mutate(across(all_of(support_cols), ~as.numeric(.)))
```

```{r overall}
# staff generally perceived internal support as fair (in the 4-5 range)
# complaint is most commonly directed at delays, followed by internal advancement and communication of expectations
support %>%
  pivot_longer(cols = all_of(support_cols), names_to = "support_area", values_to = "satisfaction") %>%
  group_by(support_area) %>%
  summarize(n = sum(!is.na(satisfaction)), 
            mean = round(mean(satisfaction, na.rm = TRUE), 2), 
            sd = round(sd(satisfaction, na.rm = TRUE), 2), # not using unbiased estimate here because sample size is sufficient
            min = min(satisfaction, na.rm = TRUE), 
            max = max(satisfaction, na.rm = TRUE)) |> 
  arrange(mean)
```

```{r by role and support category}
# develop a composite score for each major support category by weighting all constituent areas equally

# non-profit leadership reported the lowest support overall, particularly in communication and troubleshooting
# therapists reported low support in troubleshooting
support_composites <- support %>%
  rowwise() |> 
  mutate(comm_score = mean(c_across(contains("management")), na.rm = TRUE), 
         culture_score = mean(c_across(contains("team")), na.rm = TRUE), 
         help_score = mean(c_across(contains("support")), na.rm = TRUE), 
         growth_score = mean(c_across(contains("growth")), na.rm = TRUE), 
         overall_score = mean(c_across(all_of(support_cols)), na.rm = TRUE)) |> 
  ungroup()

support_composites %>%
  group_by(role) %>% 
  summarize(mean_comm = round(mean(comm_score), 2), 
            mean_culture = round(mean(culture_score), 2), 
            mean_help = round(mean(help_score), 2), 
            mean_growth = round(mean(growth_score), 2), 
            mean_overall = round(mean(overall_score), 2)) |> 
  arrange(mean_overall)
```