---
title: 'LA Crime Landscape: Analyses and Findings'
author: "Wenxin Cui"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(scales)
library(lubridate)
library(ggmap)
library(tidyverse)
```

```{r}
la_crimes_clean = read.csv('la_crimes_clean.csv')
dim(la_crimes_clean)
```

# Exploratory Visualizations

Begin the analysis by visually exploring how crime exposure differs across races.

## Contingency tables

```{r}
# the number of crimes experienced by each race across crime types
la_crimes_clean %>%
  group_by(Vict.Descent, Crm.Type) %>% 
  summarize(n = n()) %>%
  pivot_wider(names_from = Crm.Type, 
              values_from = n)
```

```{r}
# share of crimes experienced by each race across crime types
la_crimes_clean %>%
  group_by(Vict.Descent, Crm.Type) %>% 
  summarize(n = n(), .groups = 'drop_last') %>%
  mutate(prop = prop.table(n)) %>%
  select(Crm.Type, Vict.Descent, prop) %>%
  pivot_wider(names_from = Crm.Type, 
              values_from = prop)
```

## Line plots

```{r}
# each racial group's share of victims by year from 2020 to 2024
la_crimes_clean %>%
  group_by(DATE.OCC.Year, Vict.Descent) %>%
  summarize(n = n(), .groups = 'drop_last') %>%
  mutate(prop = prop.table(n)) %>%
  ggplot(aes(x = DATE.OCC.Year, y = prop, color = Vict.Descent)) + 
  geom_line() + 
  scale_y_continuous(labels = percent_format()) + 
  labs(x = "Year", y = "", color = "Victim Race") + 
  theme_minimal()
```

```{r}
# each racial group's share of victims by hour of day
la_crimes_clean %>%
  mutate(TIME.OCC.Hour = TIME.OCC %/% 100) %>%
  group_by(TIME.OCC.Hour, Vict.Descent) %>%
  summarize(n = n(), .groups = 'drop_last') %>%
  mutate(prop = prop.table(n)) %>%
  ggplot(aes(x = TIME.OCC.Hour, y = prop, color = Vict.Descent)) + 
  geom_line() + 
  scale_x_continuous(breaks = 0:23) + 
  scale_y_continuous(labels = percent_format()) + 
  labs(x = "Hour of Day", y = "", color = "Victim Race") + 
  theme_minimal()
```

## Maps

```{r}
register_stadiamaps("3e417267-0b88-4285-bd89-531d4c11168c", write = FALSE)
```

```{r}
# geographic distribution of violent crimes by race and severity
la_crimes_violent <- la_crimes_clean %>% filter(Crm.Type == "VIOLENT")
bbox <- make_bbox(LON, LAT, data = la_crimes_violent)
map <- get_stadiamap(bbox = bbox, maptype = "stamen_toner_lite", zoom = 11)
ggmap(map) + 
  geom_point(aes(LON, LAT, color = Vict.Descent), 
             data = la_crimes_violent, 
             alpha = 0.4, size = 0.1) + 
  labs(x = "Longitude", y = "Latitude", color = "Victim Race") + 
  facet_wrap(~ Part.1.2, nrow = 1, 
             labeller = labeller(Part.1.2 = c("1" = "Severe", "2" = "Less Severe"))) +
  theme_minimal()
```

```{r}
# geographic distribution of property crimes by race and severity
la_crimes_property <- la_crimes_clean %>% filter(Crm.Type == "PROPERTY")
bbox <- make_bbox(LON, LAT, data = la_crimes_property)
map <- get_stadiamap(bbox = bbox, maptype = "stamen_toner_lite", zoom = 11)
ggmap(map) + 
  geom_point(aes(LON, LAT, color = Vict.Descent), 
             data = la_crimes_property, 
             alpha = 0.4, size = 0.1) + 
  labs(x = "Longitude", y = "Latitude", color = "Victim Race") + 
  facet_wrap(~ Part.1.2, nrow = 1, 
             labeller = labeller(Part.1.2 = c("1" = "Severe", "2" = "Less Severe"))) + 
  theme_minimal()
```

```{r}
# geographic distribution of financial crimes by race and severity
la_crimes_financial <- la_crimes_clean %>% filter(Crm.Type == "FINANCIAL")
bbox <- make_bbox(LON, LAT, data = la_crimes_financial)
map <- get_stadiamap(bbox = bbox, maptype = "stamen_toner_lite", zoom = 11)
ggmap(map) + 
  geom_point(aes(LON, LAT, color = Vict.Descent), 
             data = la_crimes_financial, 
             alpha = 0.4, size = 0.1) + 
  labs(x = "Longitude", y = "Latitude", color = "Victim Race") + 
  facet_wrap(~ Part.1.2, nrow = 1, 
             labeller = labeller(Part.1.2 = c("1" = "Severe", "2" = "Less Severe"))) + 
  theme_minimal()
```

```{r}
# geographic distribution of human trafficking crimes by race and severity
la_crimes_trafficking <- la_crimes_clean %>% filter(Crm.Type == "COMMODIFIED SEX")
bbox <- make_bbox(LON, LAT, data = la_crimes_trafficking)
map <- get_stadiamap(bbox = bbox, maptype = "stamen_toner_lite", zoom = 11)
ggmap(map) + 
  geom_point(aes(LON, LAT, color = Vict.Descent), 
             data = la_crimes_trafficking, 
             alpha = 0.4, size = 0.1) + 
  labs(x = "Longitude", y = "Latitude", color = "Victim Race") + 
  facet_wrap(~ Part.1.2, nrow = 1, 
             labeller = labeller(Part.1.2 = c("1" = "Severe", "2" = "Less Severe"))) + 
  theme_minimal()
```

# Pairwise Permutation Tests

Then conduct a series of two-sample permutation tests *using total variation distance (TVD) as the test statistic* to assess the association between race and crime exposure.

```{r helper}
# helper function that computes the total variation distance between each pair of races
compute_tvds <- function(data) {
  table <- table(data$Vict.Descent, data$Crm.Type)
  prop_table <- prop.table(table, margin = 1)
  
  races <- rownames(prop_table)
  n_races <- length(races)
  
  n_pairs <- n_races * (n_races - 1) / 2
  tvds <- numeric(n_pairs) # keep track of the observed statistics
  pairs <- character(n_pairs) # keep track of the comparisons
  
  p <- 1
  for (i in 1:(n_races - 1)) {
    for (j in (i + 1):n_races) {
      pairs[p] <- paste(races[i], "vs", races[j])
      tvds[p] <- 0.5 * sum(abs(prop_table[i, ] - prop_table[j, ]))
      p <- p + 1
    }
  }
  
  names(tvds) <- pairs
  return(tvds)
}
```

```{r}
# shuffle the race labels and recompute the total variation distances 
# 1000 times to get a null distribution for each pair of races
n_perm <- 1000
n_races <- length(unique(la_crimes_clean$Vict.Descent))
n_pairs <- choose(n = n_races, k = 2)
t_null <- matrix(0, nrow = n_perm, ncol = n_pairs)

for (i in 1:n_perm) {
  la_crimes_perm <- la_crimes_clean %>%
    mutate(Vict.Descent = sample(Vict.Descent)) # sample without replacement
  t_null[i, ] <- compute_tvds(la_crimes_perm)
}

# compute the observed statistic for each pair of races
t_obs <- compute_tvds(la_crimes_clean)
colnames(t_null) <- names(t_obs)
```

```{r}
# compute the p-value for each pairwise test, using Bonferroni correction to account for multiple comparisons
alpha <- 0.05
p_vals <- sapply(1:n_pairs, function(p) {
  mean(t_null[, p] >= t_obs[p])
})
(p_vals_corrected <- p.adjust(p_vals, method = "bonferroni"))
names(p_vals_corrected) <- names(t_obs)
p_vals_corrected < alpha
```

## Visualize test results

```{r helper}
# helper function that visualizes the observed statistic against the null distribution for each pair of races
plot_statistic <- function(pair_i) {
  t_null_i <- t_null[, pair_i]
  t_obs_i <- unname(t_obs[pair_i])
  data.frame(t_null_i) %>%
    ggplot(aes(t_null_i)) + 
    geom_density(color = "black") + 
    geom_vline(xintercept = t_obs_i, color = "orange") + 
    labs(subtitle = names(t_obs[pair_i]), 
         x = "TVD", y = "Density") + 
    theme_minimal() + # place global theme here to avoid overwriting theme adjustments
    theme(plot.subtitle = element_text(hjust = 0.5))
}
```

```{r}
library(patchwork)

p1 <- plot_statistic(1)
p2 <- plot_statistic(2)
p3 <- plot_statistic(3)
p4 <- plot_statistic(4)
p5 <- plot_statistic(5)
p6 <- plot_statistic(6)

# show the plots in a 3x2 grid
(p1 + p2 + p3 + p4 + p5 + p6) + 
  patchwork::plot_layout(nrow = 3, ncol = 2, byrow = TRUE)
```