---
title: "Simulation Studies"
author: "Wenxin Cui"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

The following analyses simulate crime data to evaluate the operating characteristics of permutation tests, focusing on Type I error rate, statistical power, and the assumption of exchangeability.

# Setup

```{r data setup}
crime_types <- c("COMMODIFIED SEX", "FINANCIAL", "OTHER", "PROPERTY", "VIOLENT")
races <- c("A", "B", "H", "W")
```

```{r simulation setup}
n_sim <- 1000
n_sample <- 100
n_perm_sim <- 100
alpha <- 0.05
```

```{r test statistic setup}
# helper function that computes the total variation distance between each pair of races
compute_tvds <- function(data) {
  table <- table(data$Vict.Descent, data$Crm.Type)
  prop_table <- prop.table(table, margin = 1)
  
  races <- rownames(prop_table)
  n_races <- length(races)
  
  n_pairs <- n_races * (n_races - 1) / 2
  tvds <- numeric(n_pairs) # keep track of the observed statistics
  pairs <- character(n_pairs) # keep track of the comparisons
  
  p <- 1
  for (i in 1:(n_races - 1)) {
    for (j in (i + 1):n_races) {
      pairs[p] <- paste(races[i], "vs", races[j])
      tvds[p] <- 0.5 * sum(abs(prop_table[i, ] - prop_table[j, ]))
      p <- p + 1
    }
  }
  
  names(tvds) <- pairs
  return(tvds)
}
```

# Type I Error

Estimate the Type I error rate by repeatedly simulating data under the null hypothesis of no association between race and crime exposure, performing the designed permutation tests for each simulation, and calculating the proportion of simulations that reject the null hypothesis

```{r}
# simulate data under no association
simulate_null_data <- function(n) {
  # assume the same distribution of crime types for all races
  crime_probs <- c(0.2, 0.1, 0.3, 0.3, 0.1)
  
  # generate crimes independently of victim races
  races_sim <- sample(races, size = n, replace = TRUE)
  crimes_sim <- sample(crime_types, size = n, prob = crime_probs, replace = TRUE) # sample WITH replacement
  
  data.frame(Vict.Descent = races_sim, Crm.Type = crimes_sim)
}
```

```{r}
set.seed(411)

# run pairwise permutation tests and track the results for each simulation
pair_rejections <- matrix(0, nrow = n_sim, ncol = n_pairs)

for (s in 1:n_sim) {
  null_data <- simulate_null_data(n_sample)
  t_obs_sim <- compute_tvds(null_data)
  
  t_null_sim <- matrix(0, nrow = n_perm, ncol = n_pairs)
  for (i in 1:n_perm_sim) {
    data_perm <- null_data %>%
      mutate(Vict.Descent = sample(Vict.Descent)) # sample WITHOUT replacement
    t_null_sim[i, ] <- compute_tvds(data_perm)
  }
  
  p_vals_sim <- sapply(1:n_pairs, function(p) {
    mean(t_null_sim[, p] >= t_obs_sim[p])
  })
  pair_rejections[s, ] <- p.adjust(p_vals_sim, method = "bonferroni") < alpha
}
```

```{r}
# compute the type 1 error rate for each pairwise test
(pair_error_rates <- colMeans(pair_rejections))
```

# Statistical Power

Estimate statistical power by repeatedly simulating crime data based on victim race (i.e. there is association between race and crime exposure), performing the designed permutation tests for each simulation, and calculating the proportion of simulations that reject the null hypothesis

```{r}
# simulate data under association
simulate_alt_data <- function(n) {
  # assume the distribution of crime types differs by race
  crime_probs_by_race <- list(
    W = c(0.3, 0.2, 0.2, 0.15, 0.15),
    A = c(0.15, 0.3, 0.2, 0.2, 0.15),
    H = c(0.15, 0.4, 0.1, 0.1, 0.25),
    B = c(0.1, 0.1, 0.15, 0.15, 0.5)
  )
  
  # generate crimes based on race-specific probabilities
  races_sim <- sample(races, size = n, replace = TRUE)
  crimes_sim <- sapply(races_sim, function(race_sim) {
    sample(crime_types, size = 1, prob = crime_probs_by_race[[race_sim]])
  })
  
  data.frame(Vict.Descent = races_sim, Crm.Type = crimes_sim)
}
```

```{r}
set.seed(411)

# run pairwise permutation tests and track the results for each simulation
pair_rejections <- matrix(0, nrow = n_sim, ncol = n_pairs)

for (s in 1:n_sim) {
  alt_data <- simulate_alt_data(n_sample)
  t_obs_sim <- compute_tvds(alt_data)
  
  t_null_sim <- matrix(0, nrow = n_perm, ncol = n_pairs)
  for (i in 1:n_perm_sim) {
    data_perm <- alt_data %>%
      mutate(Vict.Descent = sample(Vict.Descent))
    t_null_sim[i, ] <- compute_tvds(data_perm)
  }
  
  p_vals_sim <- sapply(1:n_pairs, function(p) {
    mean(t_null_sim[, p] >= t_obs_sim[p])
  })
  pair_rejections[s, ] <- p.adjust(p_vals_sim, method = "bonferroni") < alpha
}
```

```{r}
# compute the power for each pairwise test
(pair_powers <- colMeans(pair_rejections))
```

# Exchangeability

Evaluate the effect of breaking exchangeability by simulating data where crime exposure and race are independent but crime types have dependencies, then evaluating the type I error rate

```{r}
# simulate crime records that are not necessarily interchangeable
simulate_dependent_null_data <- function(n) {
  races_sim <- sample(races, size = n, replace = TRUE)
  crimes_sim <- character(n)
  
  # introduce "temporal" dependency to crime observations 
  crimes_sim[1] <- sample(crime_types, size = 1)
  for (t in 2:n) {
    # assign a higher probability for the same crime for the next incident and equal probabilities for other crimes
    crime_probs <- rep(0.1, length(crime_types))
    prev_crime <- which(crime_types == crimes_sim[t - 1])
    crime_probs[prev_crime] <- 0.6
    crimes_sim[t] <- sample(crime_types, size = 1, prob = crime_probs)
  }
  
  data.frame(Vict.Descent = races_sim, Crm.Type = crimes_sim)
}
```

```{r}
set.seed(412)

# run pairwise permutation tests and track the results for each simulation
pair_rejections <- matrix(0, nrow = n_sim, ncol = n_pairs)

for (s in 1:n_sim) {
  dependent_null_data <- simulate_dependent_null_data(n_sample)
  t_obs_sim <- compute_tvds(dependent_null_data)
  
  t_null_sim <- matrix(0, nrow = n_perm, ncol = n_pairs)
  for (i in 1:n_perm_sim) {
    data_perm <- dependent_null_data %>%
      mutate(Vict.Descent = sample(Vict.Descent))
    t_null_sim[i, ] <- compute_tvds(data_perm)
  }
  
  p_vals_sim <- sapply(1:n_pairs, function(p) {
    mean(t_null_sim[, p] >= t_obs_sim[p])
  })
  pair_rejections[s, ] <- p.adjust(p_vals_sim, method = "bonferroni") < alpha
}
```

```{r}
# error rates do not differ considerably from the previous study where exchangeability holds
(pair_error_rates_dependent <- colMeans(pair_rejections))
```